Pool Model - A unified Mathematical\\ Modeling Framework for Food Science

### 2025-11-28 paper/pool_model.tex
fix supplement references
add reformulations, citations and other comments by fernando

\newcommand{\beginsupplement}{%
  \renewcommand{\thesection}{}
  \setcounter{section}{0}
  \renewcommand{\thesection}{S\arabic{section}}
  % \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}%
  % \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}%
}
\newcommand{\mbx}{\mathbf{x}}
\newcommand{\mbm}{\mathbf{m}}
\newcommand{\mbp}{\mathbf{p}}
\newcommand{\mbu}{\mathbf{u}}
\newcommand{\mbF}{\mathbf{F}}
\newcommand{\mbH}{\mathbf{H}}
\newcommand{\cmt}[1]{\textcolor{red} {[#1]}}
\newcommand{\etal}{{\textit{et al. }}}

@@

They play an especially important role within the context of food safety and production
being a key contributor to the spoilage of food products as well as fermentation processes.
Furthermore, effects such as biofilm formation can hinder drug treatment by shielding the bacterial
colony from antibiotic substances or immune response of the organism.
The natural bacterial processes are a crucial part of our planet's ecosystem, being responsible for
tasks such as nutrient cycling, decomposition of organic matter and oxygen production.

@@

During their reproductive process, bacteria undergo multiple phases~\citep{buchanan_when_1997}, starting in a lag phase which
is necessary to adapt to new environments and initially impedes the overall growth~\citep{rolfe_lag_2012, monod_growth_1949}.

@@

Widely used examples include the modified Gompertz model~\citep{Gompertz1825-wi,Gibson1987}, Baranyi-Roberts
model~\citep{baranyi_modeling_1993, baranyi_dynamic_1994} and McKellar two-compartment model~\cite{buchanan_when_1997} which distinguishes between growing and non-growing bacteria.

@@

individually-based \acp{abm} and can be effective, especially when dealing with large crowds of

@@

Bacterial growth undergoes different consecutive phases~\citep{buchanan_when_1997}, starting with the so called lag-phase, during which no increase in the bacterial count occurs.

@@

@@

The control of the population density with respect to the available nutrition by means of cell communication is called quorum sensing.
By releasing extracellular signalling molecules (autoinducers) it is possible for the bacteria to synchronize the gene expression of the whole group~\citep{ng_bacterial_2009}.
Instead of modeling complex signalling networks, we choose a simpler approach to describe the dependence of growth on the nutrient availability.

@@

The relevant question here is whether cells in the pool $D$ still contribute to the limitation of the resources by, e.g., occupying space or consuming important nutrients.

@@

It follows immediately that $\lim_{t\to\infty} G(t) = N_t-N_0\frac{\omega}{\omega+\lambda} $. The lag time $t_L$ can be approximated by the time at which half of the initial population is in the growth phase, i.e., $N_0=2G(t_L)$.

@@

Dormancy in bacteria is a reversible state where cell continue living with low metabolic activity while the cell division stops~\citep{kaprelyants_dormancy_1993}.

@@

Some examples of such mechanisms are the production of antimicrobial compounds, inhibitors or toxins~\citep{wloch-salamon_effect_2008, chao_structured_1981}.

@@

@@

@@

@@

% ================================================================================================
\newpage
\beginsupplement
% ================================================================================================
\section{General mathematical formalism}
\label{ssec:supplement1}
Consider the system consisting of $N$ species in $M$ different states.
In the general representation, all discussed in the main part systems  can be described with the Ordinary Differential Equations (\acp{ode}) using a non-linear function $\mbF$:
\begin{equation}
   \dot{\mbx} = \mbF(\mbx, \mbm, \mbp, \mbu),
\label{eq:model_ODE_x}
\end{equation}
where  $\mbx \in \mathbb{R}^{N}  \otimes \mathbb{R}^{M}$ is a vector describing bacterial count.
In matrix form it can be written as
\begin{equation}
    \mbx = \begin{pmatrix}
        x^1_1  & \dots & x^1_N  \\
        \vdots &       & \vdots \\
        x^M_1  & \dots & x^M_N  \\
            \end{pmatrix}
    \label{eq:model_bact}
\end{equation}
with $x_{j}^{i}$ describing the bacterial counts for species $i$ in state $j$.
$\mbu$ is the vector of different inputs of the model, e.g., temperature, pressure etc.
The vector $\mbm \in \mathbb{R}^{K}$ includes all variables corresponding to the microenvironment of the system which can be concentration of the inhibitors or activators:
\begin{equation}
    \dot{\mbm} = \mbH (\mbx, \mbm, \mbp, \mbu).
    \label{eq:model_microenv}
\end{equation}
Here function $\mbH$ is in general case also non-linear.\\
%
The advantage of the following formulation is its generality.
Depending on the knowledge about the system as well as the needed level of accuracy one can define an arbitrary number of states $M$ and control the complexity of the model.
Moreover, opposite to the already known Predictive Microbiology like Baranyi and Roberts one can easily change.\\
%
To calculate the bacterial count for one species $i$, we simply need to sum the tensor/vector $\mbx$ over all possible states:
\begin{equation}
    N_i = \sum_k \mbx_i^k.
\label{eq:N_pro_species}
\end{equation}
While the total bacterial count of all species of our system is
\begin{equation}
    N = \sum_i \sum_k \mbx_i^k.
\label{eq:N_total}
\end{equation}
%
%
%
Using such matrix representation of the bacterial system, we can rewrite, for example, the three-pool model with limiting resource described with equations~\ref{eq:ode_3pool_resource} in a more compact form.
The concentrations of the pools are then the components of our variable vector $\mbx (t) = (L(t), G(t), D(t))^T$.
%
\begin{align}
    \dot \mbx  &= \mbF(\mbx, \mbm, \mbp, \mbu) = \begin{pmatrix}
        -(\omega + \lambda) & 0               & 0      \\
        \lambda          & \mu R - \omega' & 0      \\
        \omega              & \omega'            & -\beta
    \end{pmatrix} \mbx\\
    \dot \mbm &= \mbH (\mbx, \mbm, \mbp, \mbu) = -\frac{\mu}{N_t} R G.
\end{align}
Separating linear and non-linear parts in two terms yields
\begin{equation}
    \dot \mbx = \begin{pmatrix}
        -(\omega + \lambda) & 0       & 0 \\
         \lambda         & \mu  & 0 \\
         \omega &  \omega'& - \beta
    \end{pmatrix} 
    \mbx + \frac{\mu}{N_t} x^2 \sum_i^3 x^i \begin{pmatrix} 0 \\ 1 \\ 0  \end{pmatrix}.
\label{eq:x_repres_3pools_resource2}
\end{equation}
Here the first term is the linear part and the second term define the quadratic behavior of the system.
%
\paragraph{Generalized Lotka-Volterra}

>>>>

\newcommand{\mbx}{\mathbf{x}}
\newcommand{\mbm}{\mathbf{m}}
\newcommand{\mbp}{\mathbf{p}}
\newcommand{\mbu}{\mathbf{u}}
\newcommand{\mbF}{\mathbf{F}}
\newcommand{\mbH}{\mathbf{H}}
\newcommand{\cmt}[1]{\textcolor{red} {[#1]}}
\newcommand{\etal}{{\textit{et al. }}}

@@

In particular, food matrices represent highly relevant microbial ecosystems, where these communities
play a key role in both food safety and spoilage as well as fermentation processes.
Biofilms are biological structures formed by complex and dynamic microbial communities adhered to
surfaces that develop different communication mechanisms (e.g. quorum sensing), and coordinate to
improve nutrient accessibility and provide protection against antimicrobial chemicals and physical
stress~\citep{Zhao2023}.
These structures are an important problem in food environments, and healthcare settings, being a
source of contamination and increased risk of infection.
% Furthermore, effects such as biofilm formation can hinder drug treatment by shielding the bacterial
% colony from antibiotic substances or immune response of the organism.
In addition, the natural bacterial processes are a crucial part of our planet's ecosystem, being
responsible for tasks such as nutrient cycling, decomposition of organic matter and oxygen production.

@@

% During their reproductive process, bacteria undergo multiple phases~\citep{buchanan_when_1997}, starting in a lag phase which
% is necessary to adapt to new environments and initially impedes the overall growth~\citep{rolfe_lag_2012, monod_growth_1949}.
During the growth process, bacteria undergo different physiological and metabolic
states~\citep{buchanan_when_1997}, that determine growth dynamics and response to the environment.
Typically, when bacteria colonize a new environment, they enter an adaptation phase, commonly
referred to as lag time~\citep{rolfe_lag_2012, monod_growth_1949}.
Under this phase, bacteria prepare the enzymatic and metabolic machinery to initiate cell division
and multiplication.
According to Baranyi and Roberts~\citep{baranyi_dynamic_1994}, lag time can
be defined as a function of the concentration of a limiting component that cells should synthesise
in order to start growing.

@@

Widely used examples include the modified Gompertz model~\citep{Zwietering1990,Gompertz1825-wi,Gibson1987}, Baranyi-Roberts
model~\citep{baranyi_modeling_1993, baranyi_dynamic_1994} and McKellar~\cite{McKellar1997} two-compartment model~\cite{buchanan_when_1997} which distinguishes between growing and non-growing bacteria.

@@

individually-based \acp{abm} and can be effective, especially when dealing with large population of

@@

Bacterial growth undergoes different consecutive phases~\citep{buchanan_when_1997}, starting with the so called lag-phase, which is a specific case of quiescence, during which no increase in the bacterial count occurs.

@@

It is a logical assumption of the speicific case of cell adaption if environmental conditions do not
change.

@@

% The control of the population density with respect to the available nutrition by means of cell communication is called quorum sensing.
We choose a minimalistic approach to describe the dependence of growth on the nutrient availability.

@@

The relevant question here is whether cells in the pool $D$ provide any significant impact on resource availability, e.g., by freeing previously occupied spaces or releasing previously consumed nutrients.

@@

It follows immediately that $\lim_{t\to\infty} G(t) = N_t-N_0\frac{\omega}{\omega+\lambda}$.
The lag time $t_L$ can be approximated by the time at which half of the initial population is in the growth phase, i.e., $N_0=2G(t_L)$.
This approach differs to other concepts of lag time such as the one by Gompertz (visual lag time) or Baranyi (work to be done).

@@

Dormancy in bacteria is a reversible state where cells continue living with low metabolic activity while the cell division stops~\citep{kaprelyants_dormancy_1993}.

@@

Some examples of such mechanisms are the production of antimicrobial compounds, inhibitors (section~\ref{section:toxin-production}) or toxins~\citep{wloch-salamon_effect_2008, chao_structured_1981}.

@@

\label{section:toxin-production}

@@

It means that ODE could still be suitable for aqueous environments, where homogeneity is more probable.

@@

By releasing these extracellular signalling molecules (autoinducers) it is possible for the bacteria to synchronize the gene expression of the whole group~\citep{ng_bacterial_2009}.

@@

% ================================================================================================
\newpage

\newcounter{suppsection}
\newcounter{supptable}
\newcounter{suppfigure}
\newcounter{suppequation}

\renewcommand{\thesection}{S\arabic{suppsection}}
\renewcommand{\thetable}{S\arabic{supptable}}%
\renewcommand{\thefigure}{S\arabic{suppfigure}}%
\renewcommand{\theequation}{S\arabic{suppequation}}%

\let\oldsection\section
\let\oldequation\equation
\let\oldtable\table
\let\oldfigure\figure
% \renewcommand{\section}{\@ifstar{\@oldsection}{\@stepcounter{suppsection}\@oldsection}}
\makeatletter
\renewcommand{\section}{\@ifstar{\oldsection*}{\stepcounter{suppsection}\oldsection}}
\renewcommand{\equation}{\@ifstar{\oldequation*}{\stepcounter{suppequation}\oldequation}}
\renewcommand{\table}{\@ifstar{\oldtable*}{\stepcounter{supptable}\oldtable}}
\renewcommand{\figure}{\@ifstar{\oldfigure*}{\stepcounter{suppfigure}\oldfigure}}
\makeatother

\addcontentsline{toc}{Section}{Supplementary Material}% Add section to ToC

% ================================================================================================
\section{General mathematical formalism}
\label{ssec:supplement1}
Consider the system consisting of $N$ species in $M$ different states.
In the general representation, all discussed in the main part systems  can be described with the Ordinary Differential Equations (\acp{ode}) using a non-linear function $\mbF$:
\begin{equation}
   \dot{\mbx} = \mbF(\mbx, \mbm, \mbp, \mbu),
\label{eq:model_ODE_x}
\end{equation}
where  $\mbx \in \mathbb{R}^{N}  \otimes \mathbb{R}^{M}$ is a vector describing bacterial count.
In matrix form it can be written as
\begin{equation}
    \mbx = \begin{pmatrix}
        x^1_1  & \dots & x^1_N  \\
        \vdots &       & \vdots \\
        x^M_1  & \dots & x^M_N  \\
            \end{pmatrix}
    \label{eq:model_bact}
\end{equation}
with $x_{j}^{i}$ describing the bacterial counts for species $i$ in state $j$.
$\mbu$ is the vector of different inputs of the model, e.g., temperature, pressure etc.
The vector $\mbm \in \mathbb{R}^{K}$ includes all variables corresponding to the microenvironment of the system which can be concentration of the inhibitors or activators:
\begin{equation}
    \dot{\mbm} = \mbH (\mbx, \mbm, \mbp, \mbu).
    \label{eq:model_microenv}
\end{equation}
Here function $\mbH$ is in general case also non-linear.\\
%
The advantage of the following formulation is its generality.
Depending on the knowledge about the system as well as the needed level of accuracy one can define an arbitrary number of states $M$ and control the complexity of the model.
Moreover, opposite to the already known Predictive Microbiology like Baranyi and Roberts one can easily change.\\
%
To calculate the bacterial count for one species $i$, we simply need to sum the tensor/vector $\mbx$ over all possible states:
\begin{equation}
    N_i = \sum_k \mbx_i^k.
\label{eq:N_pro_species}
\end{equation}
While the total bacterial count of all species of our system is
\begin{equation}
    N = \sum_i \sum_k \mbx_i^k.
\label{eq:N_total}
\end{equation}
%
%
%
Using such matrix representation of the bacterial system, we can rewrite, for example, the three-pool model with limiting resource described with equations~\ref{eq:ode_3pool_resource} in a more compact form.
The concentrations of the pools are then the components of our variable vector $\mbx (t) = (L(t), G(t), D(t))^T$.
%
\begin{align}
    \dot \mbx  &= \mbF(\mbx, \mbm, \mbp, \mbu) = \begin{pmatrix}
        -(\omega + \lambda) & 0               & 0      \\
        \lambda          & \mu R - \omega' & 0      \\
        \omega              & \omega'            & -\beta
    \end{pmatrix} \mbx\\
    \dot \mbm &= \mbH (\mbx, \mbm, \mbp, \mbu) = -\frac{\mu}{N_t} R G.
\end{align}
Separating linear and non-linear parts in two terms yields
\begin{equation}
    \dot \mbx = \begin{pmatrix}
        -(\omega + \lambda) & 0       & 0 \\
         \lambda         & \mu  & 0 \\
         \omega &  \omega'& - \beta
    \end{pmatrix} 
    \mbx + \frac{\mu}{N_t} x^2 \sum_i^3 x^i \begin{pmatrix} 0 \\ 1 \\ 0  \end{pmatrix}.
\label{eq:x_repres_3pools_resource2}
\end{equation}
Here the first term is the linear part and the second term define the quadratic behavior of the system.
%
\paragraph{Generalized Lotka-Volterra}
